---
title: "Project"
author: "Luke Hardin"
format: pdf
---

```{r}
library(tidyverse)
library(tidymodels)
library(broom)
library(dplyr)
library(Stat2Data)
library(MASS)

esg <- read.csv("SP 500 ESG Risk Ratings.csv")

```

## Introduction

One of the largest problems we as a country - and as a world - are currently facing is the climate crisis. As global temperatures continue to rise and the government sometimes tries to fight against the emission of greenhouse gases, this crisis becomes both an environmental and societal problem. Leading contributors to this environmental problem are mainly comprised of large corporations both within and outside of the United States. As environmental risks worsen day by day, it is important to address those who contribute heavily to this problem. In this study, we will examine corporations and companies belonging to the S&P 500 Index. Also known as the Standard and Poor's 500 Index, the index is market-capitalization-weighted index of 500 leading publicly traded companies in the United States (Investopedia). The S&P 500 index is regarded as one of the best gauges of prominent American equities' performance.

## Data Description

This dataset shows

## Methodology

```{r}
install.packages("naniar")
library(naniar)
vis_miss(esg)
gg_miss_fct(x = esg, fct = hyp)


```

```{r}
head(esg)

```

## Results

## Discussion
